Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic

### 2021-10-06 psy_peds_3.Rmd
updating document layout

---
title: "Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic"
output:
  html_document:
    code_folding: show
    toc: yes
    toc_float: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

# R session information

@@

@@

# Inclusion criteria

The inclusion criteria for this study is:

@@

# Variables that need to be checked/modified by each site

@@

@@

# Data input

@@

## Read csv files

### 4CE phase 2.2 files

@@

## Applying inclusion criteria to filter and format the data

@@

## First estimation: total number of patients without filtering by any disease subtype

### New hospitalizations

### Filtering on age at hospitalization time

Keeping only patients with age between 11 and 17

@@

### New admissions age range 11-17

### Total hospitalize patients age range 11-17

## Total number of patients filtering by psychiatric disorders

## Creating the inputTable

Creation of the inputTable that will be the only table used throughout the rest of the analysis - selecting only visits with at least one ICD
codes pertaining to the mental health categories - filter visits where patients are between 11 and 17 years old - add the demographic and
hospitalization related information - add the ICD code description and the related disorder group

@@

### Total patients hospitalized comparing psychiatric disorders vs other

@@

##### Percentage 
```{r}
thpd_psyPer <- count_icd %>%
  ggplot(aes(x = time_p, y = percentage_psy, fill=period)) +
  scale_fill_manual(values = cbPalette[c(3,4)]) +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  labs(y = "Percentage",
       x = paste0("Date (by ", time_period,")"),
       title = paste0("Percentage of patients with mental health related ICD codes (per ", time_period,")")) + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)+
  geom_point()
thpd_psyPer
```
##### Ratio: patients with psychiatric condition / patients without psychiatric condition

### Total patients hospitalized comparing psychiatric disorders vs other

@@

## Interrupted Time-Series Analysis

```{r}
# Model without seasonality
time_series_analysis_plotting <- function(model, df_count_psy, count_col, clearance_period = FALSE) {
  output_model <- broom::tidy(model)
  coeff_slope <- round(output_model[output_model$term == "periodduring_pandemic:time", "estimate"][[1]], 3)
  pvalue_slope <- round(output_model[output_model$term == "periodduring_pandemic:time", "p.value"][[1]], 3)
  annotation <- paste0("Lockdown slope: ", coeff_slope, " (pvalue: ", pvalue_slope, ")")
  pred_model <- predict(model, patient_count_monthly_psy, type = "response", interval = "confidence") %>% as.data.frame()
  pred_model_b <- predict(model, mutate(patient_count_monthly_psy, period = "before_pandemic"), type = "response", interval = "confidence") %>%
    as.data.frame()
  names(pred_model_b) <- paste0(names(pred_model_b), "_wo_intervention")
  pred_model <- bind_cols(df_count_psy, pred_model, pred_model_b)
  if (clearance_period) {
    pred_model$clearance_period <- ifelse(pred_model$time_p < as.Date(cut(start_clear_period, breaks = time_period)) ,
                                          "before_pandemic",
                                          ifelse(pred_model$time_p > as.Date(cut(end_clear_period, breaks = time_period)),
                                                 "during_pandemic",
                                                 "during_clearance")
    )
    pred_model$lwr <- ifelse(pred_model$clearance_period == "during_clearance", NA, pred_model$lwr)
    pred_model$upr <- ifelse(pred_model$clearance_period == "during_clearance", NA, pred_model$upr)
    pred_model$fit_wo_intervention <- ifelse(pred_model$time_p >= max(pred_model$time_p[pred_model$clearance_period == "before pandemic"]),
                                             pred_model$fit_wo_intervention,
                                             NA)
  } else {
    pred_model$clearance_period <- pred_model$period
  }
  print(pred_model$clearance_period)
  plot <- ggplot(pred_model, aes_string(x = "time_p", y = count_col)) +
    geom_point(aes(colour = clearance_period)) +
    geom_line(aes(x = time_p, y = fit_wo_intervention, colour = "Expected mean w/o pandemic"), linetype = "dashed") +
    geom_line(aes(x = time_p, y = fit, colour = clearance_period)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr, colour = clearance_period), alpha = 0.3, colour = NA) +
    scale_color_manual(name = "Time Series projection", values = c("before_pandemic" = "black",
                                                                   "Expected mean w/o pandemic" = "red",
                                                                   "during_pandemic" = "black",
                                                                   "during_clearance" = "white")) +
    annotate("text", x = as.Date("2019-10-01"), y = max(pred_model[[count_col]]) + 5, label = annotation) +
    theme(legend.position = "bottom")

@@

```{r}
time_series_analysis_plotting(ts_linear_count,
                              patient_count_monthly_psy, "count_psy",
                              clearance_period = FALSE) +
  labs(title = "Interrupted Time-Series Analysis: Percentage of encounters with psychiatric ICD codes",
       subtitle = "Without seasonality effect",
       x = time_period,
       y = "Count")
```

@@

@@

@@

#### Psychiatric conditions aggregating by suptype

##### Total hospitalizations

@@

Percentage using as denominator total number of patients

@@

### Aggregated by sex

@@

patients_hospitalized_agg_sex %>%
  ggplot(aes(x = time_p, y = count_sex, fill = sex)) +
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = time_period,
       title = paste0("Patient counts per ", time_period, " aggregated by sex"),
       subtitle = "Per Sex") + 
  theme(legend.position = "bottom") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)

##percentage
total_patients_psy <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5))

@@

## Sex
```{r}
sex <- all_codes_psy %>%
  dplyr::distinct(patient_num, sex) %>% 
  dplyr::group_by( sex ) %>%
  dplyr::summarise( count = n_distinct(patient_num) ) %>%
  dplyr::mutate( count = ifelse( count > obfuscation, count, obfuscation * 0.5),
                 total = sum( count ), percentage = round( count/total*100, 2)  )

@@

## Diagnosis

### By ICD code
```{r}
summary_diagnosis <- all_codes_psy %>%
  dplyr::distinct( patient_num, description, period, weeks, concept_code ) %>%
  dplyr::group_by( concept_code, description, period ) %>%
  dplyr::count( patient_num ) %>%
  dplyr::tally(name = "count") %>%
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::arrange( desc(count) ) %>%
  dplyr::ungroup()

@@

### By disorder group
```{r}
summary_disorder_group <- all_codes_psy %>%
  dplyr::group_by( disorder_group ) %>%
  dplyr::count( patient_num ) %>%
  dplyr::tally(name = "count") %>%
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::arrange( desc( count ))

summary_disorder_group_by_period <- all_codes_psy %>%
  dplyr::distinct( patient_num, disorder_group, period, weeks ) %>%
  dplyr::group_by( period, disorder_group ) %>%
  summarise(count = n_distinct(patient_num)) %>%
  group_by(period, add=TRUE) %>%
  mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5),
         percentage = count/sum( count )*100)

>>>>

---
title: "Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
editor_options: 
  markdown: 
    wrap: 72
---

# Setting up the R environment

### Installing packages and loading the library

@@

### R session information

@@

### Plots aesthetics information

@@

# Data retrieval

## Inclusion criteria
The inclusion criteria for this study is:

@@

## Variables that need to be checked/modified by each site

@@

5. "data_update_date": date at which the data has been updated in the local 
datawarehouse. Used to estimate patient age at time of visit, since patients age
in the 4CE demographic file is expected the age at data update.

6. "time_period": Time unit for the different graphs of the plots: months (default), weeks, days

@@

## Data loading

@@

### Reading 4CE phase 2.2 files

@@

### Applying inclusion criteria to filter and format the data

@@

### Filtering on age 11-17

@@

@@

# Results

## Admission

First calculation: total number of patients, matching the inclusion criteria, admitted on one day (without filtering by any disease subtype). Summed up over one week, or one month depending on the time unit chosen. Not filtered on qge.

## New admissions age range 11-17

## Total hospitalize patients age range 11-17

Calculating the total number of patients hospitalized on a given day (taking into account the hospitalization length).

## Patient count filtered on psychiatric disorders

### New admissions

@@

### Percentage 
```{r}
thpd_psyPer <- count_icd %>%
  ggplot(aes(x = time_p, y = percentage_psy, fill=period)) +
  scale_fill_manual(values = cbPalette[c(3,4)]) +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  labs(y = "Percentage",
       x = paste0("Date (by ", time_period,")"),
       title = paste0("Percentage of patients with mental health related ICD codes (per ", time_period,")")) + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)+
  geom_point()
thpd_psyPer
```
### Ratio: patients with psychiatric condition / patients without psychiatric condition

@@

## Interrupted Time-Series Analyses

```{r}
# Model without seasonality
time_series_analysis_plotting <- function(model, df_count_psy, count_col, clearance_period = FALSE) {
  output_model <- broom::tidy(model)
  coeff_slope <- round(output_model[output_model$term == "periodduring_pandemic:time", "estimate"][[1]], 3)
  pvalue_slope <- round(output_model[output_model$term == "periodduring_pandemic:time", "p.value"][[1]], 3)
  annotation <- paste0("Lockdown slope: ", coeff_slope, " (pvalue: ", pvalue_slope, ")")
  pred_model <- predict(model, patient_count_monthly_psy, type = "response", interval = "confidence") %>% as.data.frame()
  pred_model_b <- predict(model, mutate(patient_count_monthly_psy, period = "before_pandemic"), type = "response", interval = "confidence") %>%
    as.data.frame()
  names(pred_model_b) <- paste0(names(pred_model_b), "_wo_intervention")
  pred_model <- bind_cols(df_count_psy, pred_model, pred_model_b)
  if (clearance_period) {
    pred_model$clearance_period <- ifelse(pred_model$time_p < as.Date(cut(start_clear_period, breaks = time_period)) ,
                                          "before_pandemic",
                                          ifelse(pred_model$time_p > as.Date(cut(end_clear_period, breaks = time_period)),
                                                 "during_pandemic",
                                                 "during_clearance")
    )
    pred_model$lwr <- ifelse(pred_model$clearance_period == "during_clearance", NA, pred_model$lwr)
    pred_model$upr <- ifelse(pred_model$clearance_period == "during_clearance", NA, pred_model$upr)
    pred_model$fit_wo_intervention <- ifelse(pred_model$time_p >= max(pred_model$time_p[pred_model$clearance_period == "before pandemic"]),
                                             pred_model$fit_wo_intervention,
                                             NA)
  } else {
    pred_model$clearance_period <- pred_model$period
  }
  plot <- ggplot(pred_model, aes_string(x = "time_p", y = count_col)) +
    geom_point(aes(colour = clearance_period)) +
    geom_line(aes(x = time_p, y = fit_wo_intervention, colour = "Expected mean w/o pandemic"), linetype = "dashed") +
    geom_line(aes(x = time_p, y = fit, colour = clearance_period)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr, colour = clearance_period), alpha = 0.3, colour = NA) +
    scale_color_manual(name = "Time Series projection", values = c("before_pandemic" = "black",
                                                                   "Expected mean w/o pandemic" = "red",
                                                                   "during_pandemic" = "black",
                                                                   "during_clearance" = "white")) +
    annotate("text", x = as.Date("2019-10-01"), y = max(pred_model[[count_col]]) + 5, label = annotation) +
    theme(legend.position = "bottom")

@@

### ITS: no clearance period, no seasonality effect

```{r}
time_series_analysis_plotting(ts_linear_count,
                              patient_count_monthly_psy, "count_psy",
                              clearance_period = FALSE) +
  labs(title = "Interrupted Time-Series Analysis: Count of encounters with psychiatric ICD codes",
       subtitle = "Without seasonality effect",
       x = time_period,
       y = "Count")
```

### ITS: clearance period, no seasonality effect

@@

### ITS: no clearance period, with seasonality effect

@@

### ITS: with clearance period, with seasonality effect

@@

## Psychiatric conditions aggregating by subcategories

### New admissions: count

@@

### New admissions: percentage

@@

## Psychiatric conditions, dichotomized on patients' sex

### Count

@@

patients_hospitalized_agg_sex %>%
  ggplot(aes(x = time_p, y = count_sex, fill = sex)) +
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = time_period,
       title = paste0("Patient counts per ", time_period, " aggregated by sex"),
       subtitle = "Per Sex") + 
  theme(legend.position = "bottom") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)
```

### Percentage

```{r}
##percentage
total_patients_psy <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5))

@@

```{r}
sex <- all_codes_psy %>%
  dplyr::distinct(patient_num, sex) %>% 
  dplyr::group_by( sex ) %>%
  dplyr::summarise( count = n_distinct(patient_num) ) %>%
  dplyr::mutate( count = ifelse( count > obfuscation, count, obfuscation * 0.5),
                 total = sum( count ), percentage = round( count/total*100, 2)  )

@@

```{r}
summary_diagnosis <- all_codes_psy %>%
  dplyr::distinct( patient_num, description, period, weeks, concept_code ) %>%
  dplyr::group_by( concept_code, description, period ) %>%
  dplyr::count( patient_num ) %>%
  dplyr::tally(name = "count") %>%
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::arrange( desc(count) ) %>%
  dplyr::ungroup()

@@

```{r}
summary_disorder_group <- all_codes_psy %>%
  dplyr::group_by( disorder_group ) %>%
  dplyr::count( patient_num ) %>%
  dplyr::tally(name = "count") %>%
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::arrange( desc( count ))

summary_disorder_group_by_period <- all_codes_psy %>%
  dplyr::distinct( patient_num, disorder_group, period, weeks ) %>%
  dplyr::group_by( period, disorder_group ) %>%
  summarise(count = n_distinct(patient_num)) %>%
  group_by(period, .add=TRUE) %>%
  mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5),
         percentage = count/sum( count )*100)
